SNB Results Paper 2

### 2015-07-27 paper-1/main.tex
Almost done with editor minor comments.

In the Simon 2-stage Phase II clinical trial [\ref{Simon 1989}], a small 
number of patients are enrolled onto a therapeutic 
protocol for an initial stage. Fore each patient, we assess either success 
or failure according to a 
clinically useful criteria. Success might include tumor response to treatment
or perhaps that the response is maintained for a minimum specified duration.
In addition to failing to achieve the primary endpoint, treatment failure 
might include adverse reaction to the drug or withdrawal due to toxicity. We 
use a large number of these early successes as evidence for whether to enroll
additional patients for a second stage. A statistically small number of 
successes in Stage I would lead us to terminate the trial without entering 
Stage II.

@@

Let $I_p(s, t)$ be the {\em regularized incomplete beta function}, which is 
also one minus the c.d.f. of the negative binomial distribution. This
function satisfies $I_p(s, t) = I_{1-p}(t, s)$.  Then 
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-I_p(s, t) + 1 - I_{1-p}(t, s) \\
   &= 1. 
\end{align*}
\end{proof}
Proving that \label{eqn:pmf} is a valid probability mass function.

From (\ref{eqn:pmf}) one can see that the mass at any point on the support of the distribution may come from either $X_k = s$, captured in the function $S$, or $X_k = k-t$, captured in the function $T$. The support of $S$ is defined by $s \leq S \leq s+t+1$ and the support of $T$ is defined by $t \leq T \leq s+t+1$.

@@

Returning to the breast cancer trial introduced in the first section, a response rate of 2/12 in the first stage corresponds to an individual probability of responding of $p=0.1667$. The standard deviation of this probability is $\sqrt{p(1-p)/12}=0.1076$. If we were to use the Central Limit Theorem then the 95\% confidence interval around the mean would be $[-0.0442, 0.1076]$. While this approach is problematic since the number of samples (12 in this case) is small it does indicate how much uncertainty is built into the first stage. If we consider samples from both stages then the standard deviation drops to 0.0630 and the 95\% confidence interval would be $[0.0432, 0.2901]$.

These calculations expose two difficulties with the Simon Two-stage Optimal Design. First, success in the first stage is not a good indicator of success in the second stage and at the same time failure in first stage in no way implies failure in the second stage, especially when $p$ is close to it's threshold value. The variance in the estimate so large that it is a poor estimator. The second, difficulty is that the design is not {\em enrollment efficient}. That is, if the first stage is successful then additional patients are enrolled. Enrollment, especially for certain types of cancer, can be very difficult and it can taken time to find eligible patients. Furthermore, if all 12 of the patients in the first stage respond positively, should we really need to enroll additional patients?

Properly quantifying uncertainty in the process implies quantifying uncertainty in $p$, which stems from two sources. First is the number of samples we receive. Any estimate of $p$ will be based on data and generally the more data we have, the better our estimate. If the data are small, as they often are in clinical trials, the less reliable our estimate is. The second source comes from heterogeneity in the data. While models may assume that the probability of success $p$ is a single value this is often not the case in real-world data. In clinical trials for example an individuals chance of having an adverse outcome may be influenced by any number of underlying factors, which may vary across individual. In these cases $p$ may not be a single numeric value and may be better conceptualized as a realization over an entire distribution. A frequentest approach will not capture this variation.

@@

Both of the problems in the Simon Two-stage Optimal Design stem from the fact that it fails to quantify the uncertainty in the estimate of the success probability. A framework based on Stopped Negative Binomial distribution addresses this failure and provides the added features of monitoring the trial as new results are received, providing estimates of how many new enrollees are needed to complete a trial, and quantifying the uncertainty in the success probability.

Let's start by recasting the breast cancer study in the context of the Stopped Negative Binomial framework, assuming sequential enrollment. The first stage will stop when either two positive responses are reached or 11 negative responses. This corresponds to a Stopped Negative Binomial Distribution with $s=2$ and $t=11$. Next, lets assume that, in the first stage, there were two successes. A second success could occur during the second, through a tenth enrollment. At each of these times we would like to know the estimate for $p$ along with the lower five percent quantile of the distribution.

\begin{figure}[ht]
\includegraphics[width=\textwidth]{uncertainty.pdf}
\caption{
A visualization of the mode and the 0.05\% quantile of the estimate of the distribution of $p$.
}
\label{fig:simon}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|} \hline
{\bf Steps to 2 successes} & {\bf Mode of $p$} & {\bf 5\% quantile} \\ \hline
2 & 0.7500  & 0.4307 \\ \hline
3 & 0.5000  & 0.2355 \\ \hline
4 & 0.3000  & 0.1278 \\ \hline
5 & 0.1875  & 0.0763 \\ \hline
6 & 0.1250  & 0.0497 \\ \hline
7 & 0.0882  & 0.0347 \\ \hline
8 & 0.0652  & 0.0254 \\ \hline
9 & 0.0500  & 0.0194 \\ \hline
10 & 0.0398 & 0.0153 \\ \hline
11 & 0.0319 & 0.0123 \\ \hline
\end{tabular} 
\end{center}
\caption{
A table of the mode and the 0.05\% quantile of the estimate of the distribution of $p$.
}
\label{tab:simon}
\end{table}

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{hypo_traj.pdf}
\caption{
A visualization of the mode and the 0.05\% quantile of the estimate of the distribution of $p$ for a hypothetical sequence of patient enrollments.
}
\label{fig:hypo}
\end{figure}

Consider the trial described above. Suppose eight patients were enrolled and there was a single success on the third trial (visualized in Fig. \ref{fig:hypo}). Since the mode of the beta distribution can only be calculated when there is at least one success the maximum of the mean and the mode of the distribution is shown as the estimate. At the third time step there is an increase in the estimate and the 0.05 \% quantile corresponding to the success.

@@

Fig. \ref{fig:conditional_snb} shows the conditional, compound SNB distribution. In each of the following enrollment a success would end the trial. If none of the following enrollments are success then the trial will stop at the eleventh patient. From Fig. \ref{fig:hypo} we know that the estimate of $p$ is about 0.19. The conditional compound distribution gives the probability of another success as 0.3883 and the probability of no more success (and the first stage failing) as 0.6117. Furthermore, we can see that the probability that the trial ends in the next step is approximately 20\% and the probability the trials end with the $11^{th}$ patient is about 65\%.

@@

Suppose that in our example trial that a second successful outcome is reached on the $10^th$ patient. From the data collected we have shown how to create an estimate for the distribution of $p$. Based on that estimate we may like to see the probability of a different outcome. That is, what is the probability that we don't get two successful outcomes based on the data we have collected and what is the posterior distribution for the number of patients enrolled for the study?

In this case the success probability $p$ is distributed as $Beta(2.5, 8.8)$. Using the mode a point estimate yields $\hat{p} = 0.1667$. The standard deviation of the estimate is 0.1210. The probability of two successes during the trial is 0.6843 and the probability of failure is 0.3157. The distribution of the stopping times, along with contributions from two positive events and less than two positive events is shown below.

>>>>

In the Simon 2-stage Phase II clinical trial [\ref{Simon 1989}], a small 
number of patients are enrolled onto a therapeutic 
protocol for an initial stage. Fore each patient, we assess either success 
or failure according to a 
clinically useful criteria. Success might include tumor response to treatment
or perhaps that the response is maintained for a minimum specified duration.
In addition to failing to achieve the primary endpoint, treatment failure 
might include adverse reaction to the drug or withdrawal due to toxicity. We 
use a large number of these early successes as evidence for whether to treat
additional patients in a second stage. A statistically small number of 
successes in Stage I would lead us to terminate the trial without entering 
Stage II.

@@

Let $I_p(s, t)$ be the {\em regularized incomplete beta function}, which is 
also one minus the c.d.f. of the negative binomial distribution. This
function satisfies $I_p(s, t) = 1-I_{1-p}(t, s)$.  Then 
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-I_p(s, t) + 1 - I_{1-p}(t, s) \\
   &= 1. 
\end{align*}
\end{proof}
Proving that \label{eqn:pmf} is a valid probability mass function.

From (\ref{eqn:pmf}) one can see that the mass at any point on the support 
of the distribution may come from either $X_k = s$, captured in the 
function $S$, or $X_k = k-t$, captured in the function $T$. The support of 
$S$ is defined by $s \leq S \leq s+t-1$ and the support of $T$ is defined 
by $t \leq T \leq s+t-1$.

@@

Returning to the breast cancer trial introduced in the first section, a 
response rate of 2/12 in the first stage corresponds to an individual 
probability of responding of $p=0.1667$. The standard deviation of this 
probability is $\sqrt{p(1-p)/12}=0.1076$. If we were to use the Central 
Limit Theorem then the 95\% confidence interval around the mean would 
be $[-0.0442, 0.3776]$. While this approach is problematic since the number 
of samples (12 in this case) is small it does indicate how much uncertainty 
is built into the first stage. If we consider samples from both stages then 
the standard deviation drops to 0.0630 and the 95\% confidence interval 
would be $[0.0432, 0.2901]$.

These calculations expose two difficulties with the Simon Two-stage Optimal 
Design. First, success in the first stage is not a good indicator of success 
in the second stage and at the same time failure in first stage in no way 
implies failure in the second stage, especially when $p$ is close to it's 
threshold value. The variance in the estimate is so large that it is a poor 
estimator. The second difficulty is that the design is not {\em enrollment 
efficient}. That is, if the first stage is successful then additional patients 
are enrolled. Enrollment, especially for certain types of cancer, can be very 
difficult and can take time to find eligible patients. Furthermore, if 
all 12 of the patients in the first stage respond positively, should we 
really need to enroll additional patients?

Properly quantifying uncertainty in the process implies quantifying 
uncertainty in $p$, which stems from two sources. First is the number of 
samples we receive. An estimate of $p$ will be based on data and generally 
the more data we have, the better our estimate. If the data are small, as 
they often are in clinical trials, the less reliable our estimate is. The 
second source comes from heterogeneity in the data. While models may assume 
that the probability of success $p$ is a single value this is often not the 
case in real-world data. In clinical trials for example an individual's chance 
of having an adverse outcome may be influenced by any number of underlying 
factors, which may vary across individual. In these cases $p$ may not be a 
single numeric value and may be better conceptualized as a realization over 
an entire distribution. A frequentest approach will not capture this variation.

@@

Both of the problems in the Simon Two-stage Optimal Design stem from the fact that it fails to quantify the uncertainty in the estimate of the success probability. A framework based on the SNB distribution addresses this failure and 
provides the added features of monitoring the trial as new results are 
received, providing estimates of how many new enrollees are needed to complete 
a trial, and quantifying the uncertainty in the success probability.

Let's start by recasting the breast cancer study in the context of the SNB 
framework, assuming sequential enrollment. The first stage will stop when 
either two positive responses are reached or 11 negative responses. This 
corresponds to a SNB distribution with $s=2$ and $t=11$. Next, let's assume 
that, in the first stage, there were two successes. A second success could 
occur during the second, through a tenth enrollment. At each of these times we 
would like to know the estimate for $p$ along with the lower five percent 
quantile of the distribution.

\begin{figure}[ht]
\includegraphics[width=\textwidth]{uncertainty.pdf}
\caption{
A visualization of the mode and the 5\% quantile of the estimate of the distribution of $p$.
}
\label{fig:simon}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|} \hline
{\bf Steps to 2 successes} & {\bf Mode of $p$} & {\bf 5\% quantile} \\ \hline
2 & 0.7500  & 0.4307 \\ \hline
3 & 0.5000  & 0.2355 \\ \hline
4 & 0.3000  & 0.1278 \\ \hline
5 & 0.1875  & 0.0763 \\ \hline
6 & 0.1250  & 0.0497 \\ \hline
7 & 0.0882  & 0.0347 \\ \hline
8 & 0.0652  & 0.0254 \\ \hline
9 & 0.0500  & 0.0194 \\ \hline
10 & 0.0398 & 0.0153 \\ \hline
11 & 0.0319 & 0.0123 \\ \hline
\end{tabular} 
\end{center}
\caption{
A table of the mode and the 5\% quantile of the estimate of the distribution of $p$.
}
\label{tab:simon}
\end{table}

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{hypo_traj.pdf}
\caption{
A visualization of the mode and the 5\% quantile of the estimate of the distribution of $p$ for a hypothetical sequence of patient enrollments.
}
\label{fig:hypo}
\end{figure}

Consider the trial described above. Suppose eight patients were enrolled and 
there was a single success on the third trial (visualized in Fig. 
\ref{fig:hypo}). The mode of the beta distribution can only be calculated when 
there is at least one success. The mean of the distribution can be
taken otherwise. At the third time step there is an increase in the estimate and the 5\% quantile corresponding to the success.

@@

Fig. \ref{fig:conditional_snb} shows the compound SNB distribution. In each of the following enrollment a success would end the trial. If none of the following enrollments are success then the trial will stop at the eleventh patient. From Fig. \ref{fig:hypo} we know that the estimate of $p$ is about 0.19. The compound distribution gives the probability of another success as 0.3883 and the probability of no more success (and the first stage failing) as 0.6117. Furthermore, we can see that the probability that the trial ends in the next step is approximately 20\% and the probability the trials end with the $11^{th}$ patient is about 65\%.

@@

Suppose that, in our example trial, a second successful outcome was reached 
on the $10^{th}$ patient. From the data collected we have shown how to create 
an estimate for the distribution of $p$. Based on that estimate we may like to 
see the probability of a different outcome. We may want to answer the
question, "What is the probability that we don't get two successful outcomes?" 
We may also ask what is the distribution for the number of patients 
enrolled for a similar study?

If there are two successful enrollments with the second one occurring 
on the $10^{th}$ trial the success probability $p$ is distributed as 
$Beta(2.5, 8.8)$. 
Using the mode, a point estimate yields $\hat{p} = 0.1667$. The standard 
deviation of the estimate is 0.1210. The probability of two successes during 
the trial is 0.6843 and the probability of failure is 0.3157. The distribution 
of the stopping times, along with contributions from two positive events and 
less than two positive events is shown below.

### 2015-07-27 paper-2/main.tex
Almost done with editor minor comments.

\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{prop}{Proposition}

@@

\author[1]{Michael J. Kane}
\author[1]{Daniel Zelterman}

\affil[1]{Department of Biostatistics, Yale University, New Haven CT, USA}

@@

\begin{abstract}
This paper other properties of the snb...
\end{abstract}

\section{Introduce the Z-Plot Construction}

\begin{figure}[ht]
\includegraphics[width=\textwidth]{z-plot1.pdf}
\caption{
A hypothetical realization of the Bernoulli process.
}
\label{fig:z_plot}
\end{figure}

\section{Rotating the Z-Plot}

>>>>

\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{prop}{Proposition}
\newtheorem{thm}{Theorem}

@@

\author{Michael J. Kane}
\author{Daniel Zelterman}

\affil{Department of Biostatistics, Yale University, New Haven CT, USA}

@@

%\begin{abstract}
%This paper other properties of the snb...
%\end{abstract}

\section{The Stopped Negative Binomial Distribution} \label{sect:intro}

\begin{enumerate}
\item Review of the SNB
\item Show when the SNB is and isn't unimodel.
\item Past efforts to show the asymptotic behavior using Sterling 
approximations.
\end{enumerate}

\section{Main Proof}

\begin{thm} \label{thm:main}
The main theorem states that once there is a central limit effect 
then the hitting times look the L\'{e}vy-Bachlier formula for the density 
of the first hitting time of a sloped line.
\end{thm}

The proof of Theorem \ref{thm:main} proceeds as follows. In Section
\ref{sect:process_construction} it is shown that the SNB can be 
constructed as a binomial process on a grid. Under suitable 
rotation and scaling the process is equivalent to a random walk 
on the integers with stopping rules. Section \ref{sect:fclt} shows that
after reparameterizing the process it converges to a Standard Brownian
Motion with stopping rules. The density of the first hitting times for
the two separate endpoints are each distributed according to the 
L\'{e}vy-Bachlier formula. Section \ref{sect:stat} provides statistical 
results, deriving the convergence when the rotation and scaling parameters
are estimated from data. Section \ref{sect:discussion} ends the article
with a discussion and proposes avenues for future work.

\section{Process Construction} \label{sect:process_construction}

As shown in Section \ref{sect:intro} the SNB can be conceptualized as
a series of coin flips that stop when either $s$ heads or $t$ tail are
reached. The time to reach either $s$ or $t$ is distributed as the SNB.
The process can be visualized as in Figure \ref{z-plot1.pdf}. The process
starts at the origin and at each step a coin is flipped if it is heads
the process moves in the vertical direction one step, if tails it moves
one step in the horizontal direction.

\begin{figure}[ht]
\includegraphics[width=\textwidth]{z-plot1.pdf}
\caption{
A hypothetical realization of the SNB coin-flipping process.
}
\label{fig:z_plot}
\end{figure}

The expected slope of the process is determined by $p \in [0, 1]$, the 
probability of a heads. If $s$ and $t$ are finite, then as $p$ increases 
the probability a given realization of the process hits $s$ increases (and
$t$ decreases). Likewise as the $p$ decreases the probability a realization 
hits $s$ decreases (and $t$ increases).

Letting $p = \alpha / \alpha + \beta$, the expected direction of the 
process is $\theta = \text{atan}(\alpha / \beta)$. In Section \ref{sect:stat}
it is shown that an estimate of $\theta$ converges at the same rate as
an estimate of $p$. For now though, we would like to rotate the process
clockwise by $\theta$ so that, in expectation, the rotate process angle
is zero.

\begin{lemma}
After rotation, the line $y=s$ becomes
\begin{equation}
y' = -\tan(\theta) x' + s \left(\sin(\theta) \tan(\theta) + \cos(\theta) \right)
\end{equation}
where $y'$ and $x'$ are the vertical and horizontal axes in the rotation
space.
\end{lemma}

\begin{enumerate}
\item Show a picture of the z-plot before and after rotation/rescaling.
\item Rotate the system clockwise by $\theta = p = \alpha/(\alpha+\beta)$ 
where $\beta$ is the number of failures and $\alpha$ is the number of successes.
\item Derive the rotated $t$- and $s$-lines.
\item Calculate the scaling for the rotated $t$- and $s$- lines so that the
reparameterized process goes right and either up on step or down one step...
Like a random walk on the integers over time.
\end{enumerate}

\section{The FCLT for the process} \label{sect:fclt}

\begin{enumerate}
\item TODO: Find the appropriate scalings for $s$ an $t$ in terms of $n$
\item Use the Heyde/Hall FCLT to show that the symmetric random walk
converges to a Standard Brownian motion over the appropriate interval.
\item Derive the probability of hitting $s$ or $t$ as a Girsinov process.
\end{enumerate}

Describe the properties of the Girsinov process and relate them back to 
the multi-mode result from the beginning.

\section{Estimating $\theta$ and the rotations from data} \label{sect:stat}

\begin{enumerate}
\item Show that if we need to estimate $\theta$ that it converges like 
$\widehat{p}$. Include the graph of $\widehat{\theta}$ vs. $\alpha$.
\item Show convergence results for for the $s$- and $t$-lines. 
\end{enumerate}

\section{Discussion and Future Work} \label{sect:discussion}
